---
title: "Homework Assignment #4: Prioritizing Potential Aquaculture"
subtitle: "Suitability Function"
author: "Carmen Hoyt"
date: last-modified
execute: 
  warning: false
  messages: false
format:
  html:
    toc: true
    code-fold: true
editor_options: 
  chunk_output_type: console
---

## Import packages
```{r}
#| code-summary: Expand code
library(here)
library(tidyverse)
library(sf)
library(tmap)
library(dplyr)
library(kableExtra)
library(terra)
library(patchwork)
library(tmaptools)
```

## Read in data
```{r}
#| code-summary: Expand code
# Read in EEZ data
west_coast_eez <- read_sf(here::here("data", "wc_regions_clean.shp"))

# Read in and rasterize bathymetry data
bathymetry <- stars::read_stars(here::here("data", "depth.tif"))
bathymetry <- rast(bathymetry)

# Read in and rasterize SST (sea surface temp) data
sst_rasters <- list.files(here("data"), pattern = glob2rx("average_annual*"), full.names = TRUE)

for (i in seq_along(sst_rasters)) {
  raster <- stars::read_stars(sst_rasters[i])
  raster <- rast(raster)
  raster_name <- paste0("sst", str_sub(sst_rasters[i], -9 ,-5))
  if (st_crs(west_coast_eez) == st_crs(raster)) {
  print("The CRS match! Good to go.")
  } else {
  raster <- project(raster, crs(west_coast_eez))
  }
  try(if(st_crs(west_coast_eez) != st_crs(raster)) stop("CRS doesn't match."))
  assign(raster_name, raster)
}

# Stack SST rasters
sst <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)

# Find mean SST
avg_sst <- mean(sst, na.rm = TRUE)

# Convert from Kelvin to Celsius
avg_sst <- avg_sst - 273.15

# Clip bathymetry to same extent as SST
depth <- bathymetry[avg_sst, drop = FALSE]

# Resample depth to match SST resolution
depth_resample <- terra::resample(depth, y = avg_sst, method = "near")

# Stack SST and depth rasters
stack <- c(sst, depth_resample)
```

## Suitability Function

The function `suitability()` takes the following arguments (in order):

- `species_name`: the name of the species (as a string)

- `min_temp`: minimum sea surface temperature (degrees Celsius)

- `max_temp`: maximum sea surface temperature (degrees Celsius)

- `min_depth`: the absolute value of the shallowest preferred depth (meters)

- `max_depth`: the absolute value of the deepest preferred depth (meters)

The final product is a map of West Coast EEZ regions colored by area (km^2) associated with suitable growing conditions for the species of interest. 

Example: `suitability("Abalone", 8, 18, 0, 24)` where:

- `species_name`: "Abalone"

- `min_temp`: 8 (degrees C)

- `max_temp`: 18 (degrees C)

- `min_depth`: 0 (meters)

- `max_depth`: 24 (meters)

```{r}
#| code-summary: Expand code
# Create suitability function
suitability <- function(species_name, min_temp, max_temp, min_depth, max_depth) {
  
  # Save species name
  species_name = species_name
  
  # Reclassify temp raster for suitable temperatures
  rcl_temp <- matrix(c(-Inf, min_temp, 0, # min, max, group
               min_temp, max_temp, 1,
               max_temp, Inf, 0),
             ncol = 3, byrow = TRUE)

  sst_suitable <- terra::classify(avg_sst, rcl = rcl_temp)

  # Reclassify depth raster for suitable depths
  rcl_depth <- matrix(c(-Inf, -max_depth, 0, # min, max, group
               -max_depth, -min_depth, 1,
               -min_depth, Inf, 0),
             ncol = 3, byrow = TRUE)

  depth_suitable <- terra::classify(depth_resample, rcl = rcl_depth)

  # Stack the two reclassified temp and depth rasters and multiply to find where suitability overlaps 
  fun <- function(x,y){return(x*y)}
  stack <- c(sst_suitable, depth_suitable)
  suitable_raster <- lapp(stack, fun)

  # Replace 0 values with NA
  suitable_raster[suitable_raster == 0] <- NA
  
  # Find suitable area masked to US EEZ extent
  suitable_raster <- mask(suitable_raster, west_coast_eez)
  
  # Rasterize the EEZ to calculate area
  eez_raster <- rasterize(west_coast_eez, stack, "rgn")
  
  # Find the size of each cell in suitable_raster
  cell_size <- cellSize(suitable_raster, unit = "km")
  
  # Calculate total suitable area for each EEZ and save as a dataframe
  area <- data.frame((zonal(cell_size*suitable_raster, 
                                      z = eez_raster, 
                                      fun ="sum", 
                                      na.rm = TRUE)))
  
  # Join to EEZ vector to create geo-dataframe to map
  map_df <- left_join(west_coast_eez, area, by = 'rgn') %>%
  select('rgn', 'area', 'area_km2') %>%
  mutate(area_rounded = round(area),
         Percent = round(area/area_km2*100, 2)) %>%
  mutate(labels = paste0(rgn, ": ", "\n", area_rounded, " km^2"))
  
  # Map suitable area by EEZ region
  suitability_map <- tm_shape(avg_sst) +
    tm_raster(palette = "Greys",
            title = "Mean SST (C)") +
    tm_shape(map_df) +
    tm_polygons(title = "Suitable Area \n(km^2)",
              col = 'area_rounded',
              palette = "Purples",
              legend.show = TRUE) +
  tm_text("labels", size = 1/2) +
  tm_layout(main.title = paste("Suitable", species_name, "Habitat"),
          legend.position = c("right", "top"),
          bg.color = "white",
          legend.outside = FALSE) +
  tm_compass(position = c("left", "bottom")) +
  tm_scale_bar(position = c("left", "bottom"))
  map_name <- paste0(species_name, "_map")
  assign(map_name, suitability_map)
  
  return(suitability_map)
}
```

## Use Function
```{r}
#| code-summary: Expand code
suitability("Oyster", 11, 30, 0, 70)
suitability("Abalone", 8, 18, 0, 24) 
```

## Data Sources

| Data | Citation | Link |
|------------------|------------------|------------------|------------------|
| Species Requirements for Red Abalone | Bisby, F.A., M.A. Ruggiero, K.L. Wilson, M. Cachuela-Palacio, S.W. Kimani, Y.R. Roskov, A. Soulier-Perkins and J. van Hertum. 2005. (via SeaLifeBase) | [https://www.sealifebase.ca/summary/Haliotis-rufescens.html](https://www.sealifebase.ca/summary/Haliotis-rufescens.html) |
| Sea Surface Temperature Data | NOAA Coral Reef Watch. 2018, updated daily. Daily Global 5km Satellite Sea Surface Temperature Anomaly (Version 3.1, released August 1, 2018). College Park, Maryland, USA: NOAA Coral Reef Watch | [https://coralreefwatch.noaa.gov/product/vs/data.php](https://coralreefwatch.noaa.gov/product/vs/data.php) | 
| Bathymetry Data | GEBCO Compilation Group (2024) GEBCO 2024 Grid (doi:10.5285/1c44ce99-0a0d-5f4f-e063-7086abc0ea0f) | [https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area](https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area) |
| EEZ Boundary | MarineRegions. nd. EEZ Boundaries. | [https://www.marineregions.org/eez.php](https://www.marineregions.org/eez.php) |

